DOpElib:\\ Differential Equations and Optimization Environment

### 2018-10-17 Examples/PDE/InstatPDE/Example8/content.tex
add examples to parallel-dev

\subsubsection{General problem description}
We consider phase-field fracture propagation on a slit domain, with timedependet 
Dirichlet data and homogenous Neumann data.
The main novelties are:
\begin{itemize}
\item Imposing a variational inequality constraint ($\frac{\partial \varphi}{\partial t} \ge 0$)
via penalization;
%\item Implementation of an augmented Lagrangian loop (not yet)
\item Stress splitting into tension and compression \`a
la \cite{MieWelHof10a}, see as well \cite{AmorMarigoMaurini2009}, located 
in 
\begin{verbatim}
stress_splitting.h
\end{verbatim}
\item Implementation of a Newton solver that 
allows for an increase of the residual for which, however,
no convergence results is available:
\begin{verbatim}
DOpEsrc/templates/instat_step_modified_newtonsolver.h
\end{verbatim}
\item Two localpde* files:
\begin{verbatim}
localpde_quasi_monolithic.h
localpde_fully_implicit.h
\end{verbatim}
that provide implementations of a quasi-monolithic scheme (similar, but not
the same as \cite{HeWheWi15}) and a fully implicit implementation.
The fully implicit implementation requires a modified Newton scheme that 
allows for a temporary increase of the Newton residual \cite{Wi17_pff_error_oriented_Newton}.
\end{itemize}
Let $u$ be the displacement and $\varphi$ be the phase field variable which should highlight the crack.
Furthermore we denote the $L^2(\Omega)$ inner product of $v_1$ and $v_2$ with $(v_1,v_2)$.
A material is supposed to be undamaged at position $x$ if $\varphi(x)$ is close to 1 and comletely cracked
if it is close to $0$.
We want to find a stationary point of the energy functional
\begin{align*}
E(u, \varphi):=& \frac{1}{2}(\sigma(u,\varphi),\varepsilon(u)) \\
+&\frac{G_c}{2}(\frac{1}{\varepsilon} \Vert 1 - \varphi \Vert^2)+\varepsilon \Vert \nabla \varphi  \Vert\\
+&\underbrace{\frac{\gamma_{pen}}{2} ((\varphi(t_n)-\varphi(t_{n-1}))_+)^2}_{\text{penalization term}},
\end{align*}
with
\begin{align*}
\sigma(u,\varphi)&:= (1-\kappa)\varphi^2 \sigma(u)^+ + \sigma(u)^-,\\
\sigma(u)^+ &:= 2*\mu \varepsilon(u)^+ +\lambda (trace(\varepsilon(u)))_+ I,\\
\sigma(u)^- &:= 2*\mu \varepsilon(u)^- +\lambda (trace(\varepsilon(u)))_- I,\\
(c)_+ &:= max(0, c), \\
(c)_- &:= c-(c)_+, \\
\varepsilon(u)^+ &:= \sum_{i=1}^{d} (\lambda_i)_+ n_i^T n_i,\\
\varepsilon(u)^- &:= \sum_{i=1}^{d} (\lambda_i)_- n_i^T n_i,\\
\varepsilon(u)   &:= \frac{1}{2}(\nabla u +(\nabla u)^T),
\end{align*}
where $c \in \mathbb{R}$, $t^n$ denotes the time at timestep $n$, $\lambda_i$ denotes the $i-th$ eigenvalue of $\varepsilon(u)$ and $n_i$ the corresponding eigenvector.
The idea for the choice of the energy functional is presented
in \cite{MieHofWel2010} (based on the original work \cite{BourFraMar00,BourFraMar08,FraMar98}).
This leads to the following problem: Find $(u, \varphi) \in V_u \times V_\varphi$ such that

>>>>

\subsubsection{General problem description}
We consider phase-field fracture propagation on a slit domain, with time dependent 
Dirichlet data and homogenous Neumann data.
The main novelties are:
\begin{itemize}
\item Imposing a variational inequality constraint ($\frac{\partial \varphi}{\partial t} \leq 0$)
via penalization;
%\item Implementation of an augmented Lagrangian loop (not yet)
\item Stress splitting into tension and compression \`a
la \cite{MieWelHof10a}, see as well \cite{AmorMarigoMaurini2009}, located 
in 
\begin{verbatim}
stress_splitting.h
\end{verbatim}
\item Implementation of a Newton solver that 
allows for an increase of the residual for which, however,
no convergence results are available:
\begin{verbatim}
DOpEsrc/templates/instat_step_modified_newtonsolver.h
\end{verbatim}
\item Two localpde* files:
\begin{verbatim}
localpde_quasi_monolithic.h
localpde_fully_implicit.h
\end{verbatim}
that provide implementations of a quasi-monolithic scheme (similar, but not
the same as \cite{HeWheWi15}) and a fully implicit implementation.
The fully implicit implementation requires a modified Newton scheme (see the
previous bullet point) that 
allows for a temporary increase of the Newton residual (similar to \cite{Wi17_pff_error_oriented_Newton}).
\end{itemize}
Let $u$ be the displacement and $\varphi$ be the phase field variable which should highlight the crack.
Furthermore we denote the $L^2(\Omega)$ inner product of $v_1$ and $v_2$ with $(v_1,v_2)$.
A material is supposed to be undamaged at position $x$ if $\varphi(x)$ is close to 1 and comletely cracked
if it is close to $0$.
We want to find a stationary point of the energy functional
\begin{align*}
E(u, \varphi):=& \frac{1}{2}(\sigma(u,\varphi),\varepsilon(u)) \\
+&\frac{G_c}{2}(\frac{1}{\varepsilon} \Vert 1 - \varphi \Vert^2)+\varepsilon \Vert \nabla \varphi  \Vert\\
+&\underbrace{\frac{\gamma_{pen}}{2} ((\varphi(t_n)-\varphi(t_{n-1}))_+)^2}_{\text{penalization term}},
\end{align*}
with
\begin{align*}
\sigma(u,\varphi)&:= (1-\kappa)\varphi^2 \sigma(u)^+ + \sigma(u)^-,\\
\sigma(u)^+ &:= 2*\mu \varepsilon(u)^+ +\lambda (trace(\varepsilon(u)))_+ I,\\
\sigma(u)^- &:= 2*\mu \varepsilon(u)^- +\lambda (trace(\varepsilon(u)))_- I,\\
(c)_+ &:= max(0, c), \\
(c)_- &:= c-(c)_+, \\
\varepsilon(u)^+ &:= \sum_{i=1}^{d} (\lambda_i)_+ n_i^T n_i,\\
\varepsilon(u)^- &:= \sum_{i=1}^{d} (\lambda_i)_- n_i^T n_i,\\
\varepsilon(u)   &:= \frac{1}{2}(\nabla u +(\nabla u)^T),
\end{align*}
where $c \in \mathbb{R}$, $t^n$ denotes the time at timestep $n$, $\lambda_i$ denotes the $i-th$ eigenvalue of $\varepsilon(u)$ and $n_i$ the corresponding eigenvector.
The idea for the choice of the energy functional is presented
in \cite{MieHofWel2010} (based on the original work \cite{BourFraMar00,BourFraMar08,FraMar98}).
This leads to the following problem: Find $(u, \varphi) \in V_u \times V_\varphi$ such that

### 2018-10-17 Examples/PDE/StatPDE/Example1/content.tex
add examples to parallel-dev

@@

\vspace{0.2cm}
\index{functional evaluation}
Here, we declare all quantities of interest (functionals), e.g. point values, drag, lift, mean values of certain quantities over a subdomain etc. \\
Each of these functionals is declared as a class of its own, but in \texttt{DOpElib} all classes are derived from a so-called \texttt{FunctionalInterface} class.\\
In the current example we declare functionals for point values of the velocity and for the flux at the outflow boundary of the channel.\index{functional!point value} \index{functional!boundary flux}\\

>>>>

As code verification, we evaluate two different types of functionals. 
First a point functional measuring the $x$-velocity and 
a flux functional 
\[
\int_{\Gamma_{out}} v\cdot n \, ds,
\]
on the outflow boundary. Both a described in the 
\texttt{functionals.h} file as described below.

@@

\vspace{0.2cm}
\index{functional evaluation}
Here, we declare all quantities of interest (functionals), e.g. point values, drag, lift, mean values of certain quantities over a subdomain etc. \\
Each of these functionals is declared as a class of its own, but in \texttt{DOpElib} all classes are derived from a so-called \texttt{FunctionalInterface} class.\\
As already mentioned previsously, in the current example we declare functionals for point values of the velocity and for the flux at the outflow boundary of the channel.\index{functional!point value} \index{functional!boundary flux}\\

### 2018-10-17 Examples/PDE/StatPDE/Example13/content.tex
add examples to parallel-dev

\subsubsection{General problem description}
Within this example, we demonstrate how to use the
dG (discontinuous Galerkin) method for the solution of 
a transport equation. The example corresponds to 
the dealii step-12.

\subsubsection{Implementational Details}

>>>>

\subsubsection{General problem description}
Within this example, we demonstrate how to use the
dG (discontinuous Galerkin) method for the solution of 
a transport equation. The example corresponds to 
the dealii step-12. Here we want to solve the transport equation 
\begin{align*}
 \nabla \cdot (\beta u) &= 0 & \text{in }&\Omega,\\
 u &= g &\text{on }&\Gamma_- := \{ x \in \partial \Omega\,|\, \beta(x) \cdot n(x) < 0\}.
\end{align*}
Where $n$ is the outward unit normal, $\Omega = (0,1)^2$ and 
\[
\beta(x) = \frac{1}{|x|} \begin{pmatrix} -x_2\\x_1
  \end{pmatrix}.
\]

For the numerical solution, as in dealii step~12, we use the upwind discontinuous 
Galerkin \index{discontinuous Galerkin}. Hence we solve the problem of finding $u_h$ 
such that 
\[
 -\sum_{T \in \mathcal T_h} (u_h \beta \cdot  \nabla v_h)_T 
 + \sum_{F \in \mathcal F_h} (u_h^-,[\beta \cdot n v_h])_F + (u_h,\beta \cdot n v_h)_{\Gamma_+} = -(g,\beta \cdot n v_h)_{\Gamma_-}
\]
where $\Gamma_+ =  \{ x \in \partial \Omega\,|\, \beta(x) \cdot n(x) > 0\}$, $\mathcal T_h$ and $\mathcal F_h$ denote the 
elements and interior faces of the mesh, respectively. The jump is defined as 
\[
[\beta \cdot n v_h] = (v^+ - v^-) \beta \cdot n^+
\]
where the superscript $+$ or $-$ denotes the dependence on the upstream $+$ or downstream $-$ element.

\subsubsection{Implementational Details}
Within this program, we need to make use of the additional \texttt{Face*} and \texttt{Interface*} methods as given in the  
\texttt{PDEInterface} class. The Face* methods define all integrals on $\mathcal F$ in which the element interacts 
with it self. The Interface* methods are used for the coupling between the two neighboring elements over the 
given face.

The program requires the following changes in contrast to the prior examples:

\paragraph{main.cc} We utilize Block Preconditioners for the solution of the resulting system. 
To this end we included the line
\begin{verbatim}
typedef DOpEWrapper::PreconditionBlockSSOR_Wrapper<MATRIX,4> 
        PRECONDITIONERSSOR; 
\end{verbatim}
In contrast to all other preconditioners, we need to specify the block size. This number needs 
to correspond to the number of unknowns per elements; here 4 since we use Q1-elements. Note that this works for dG
elements only.

The next important change is that we now not only use a discontinuous element, but we will need to assemble 
terms on faces between elements that couple the unknowns in the different elements together. For this,
the matrix needs to have the corresponding non-zero entries specified in the sparsity pattern.  
To do this, the space time handler has an argument \texttt{bool flux\_pattern} in the constructor that needs to be set to true, i.e., we instantiate as follows: 
\begin{verbatim}
STH DOFH(triangulation, state_fe, true);
\end{verbatim}

Finally, the integration will utilize a special function to be declared in the \texttt{LocalPDE}, 
hence all objects must use the \texttt{LocalPDE}
and not the \texttt{PDEInterface}.
To make sure that this is the case, the \texttt{PDEProblemContainer} needs to be initialized with the following arguments
\begin{verbatim}
typedef PDEProblemContainer<LocalPDE<CDC, FDC, DOFHANDLER, VECTOR, DIM>,
    SimpleDirichletData<VECTOR, DIM>, SPARSITYPATTERN, VECTOR, DIM> OP;
\end{verbatim}

\paragraph{localpde.h}
In order to integrate the PDE above, we have to deal with one term that has not been considered before
\[
 \sum_{F \in \mathcal F_h} (u_h^-,[\beta \cdot n v_h])_F 
\]
Since internally all terms by sums over elements we split this term into contributions on the 
element edges $\partial K$. On an element $K$ a face $F$, with outward normal $n$ 
connects to another element $K'$, depending on the sign of $\beta \cdot n$ we have two cases. 
$\beta \cdot n > 0$ in which case $u^-_h = u_h$ or $\beta \cdot n < 0$ in which case $u^-_h = u_h^* = u_h\lvert_{K'}$, i.e.
the value from the neighbor. The jump always contains the values $v_h$ and $v_h^*$.

Let now $\beta \cdot n > 0$. Then we assemble the contributions coming only from this element in the \texttt{FaceEquation} 
(and \texttt{FaceMatrix}),
i.e.,
\[
 (u_h,\beta \cdot n v_h)_F
\]
The other part of the jump, namely 
\[
 -(u_h,\beta \cdot n v_h^*)_F
\]
is not assembled here, since the test functions do not live on the selected element $K$. These contributions will
be assembled once the element $K'$ is selected (and hence on the same face $\beta \cdot n < 0$.
Once this is the case, i.e. $\beta \cdot n < 0$, we assemble the other part of the jump,
which is now 
\[
(u_h^*,\beta \cdot n v_h)_F.
\]
This is done in the \texttt{InterfaceEquation} (and \texttt{InterfaceMatrix}) since we couple unknowns for the neighboring element $K^*$ (the values of $u_h^*$) with those on $K$ (the values of $v_h$). Note that in contrast to the view on the element with 
$\beta \cdot n > 0$ the term has apparently switched signs. This is no typo, but due to the fact, that the outward normal 
has changed direction.

The precise assembly is analogous to the usual integrals, hence we don't provide more details. The only 
thing different in the \texttt{InterfaceEquation} and \texttt{InterfaceMatrix} we need to access the 
values on the element on the other side of the face. To this end, all \texttt{Get*} functions used, 
such as \texttt{GetFEFaceValuesState},
have a counterpart \texttt{GetNbr*}, i.e., \texttt{GetNbrFEFaceValuesState}, to access the corresponding values 
on the neighboring element.

Naturally the two functions
\begin{verbatim}
bool HasFaces() const;
bool HasInterfaces() const;
\end{verbatim}
need to return \texttt{true}.

A last and important change is that we now need to implement the 
method 
\begin{verbatim}
template<typename ELEMENTITERATOR>
 bool AtInterface(ELEMENTITERATOR& element, unsigned int face) const
 {
   if (element[0]->neighbor_index(face) != -1) 
     return true;
   return false;
 }
\end{verbatim}
that returns true whenever we are on an interior face and false otherwise.

### 2018-10-17 Examples/PDE/StatPDE/Example3/content.tex
add examples to parallel-dev

\subsubsection{General problem description}
\index{stationary PDE}
In this example we consider the same setting as in subsection~\ref{PDE_Stat_Stokes}, the only
difference is that we want to employ the hp-Finite-Elements. So the equation we solve is still 
the stationary incompressible Stokes equation \index{stationary PDE!Stokes equation}. Here,
we use the symmetric stress tensor which has a little consequence when using 
the do-nothing outflow condition. In strong formulation we have
\begin{align} 
-\frac{1}{2}\nabla\cdot (\nabla v + \nabla v^{T}) + \nabla p &= f \\ 
\nabla \cdot v &= 0 \notag
\end{align} 
on the domain $\Omega = [-6,6]\times [0,2]$.  We split $\partial \Omega = \Gamma_D \cup \Gamma_{out}$. The right hand side of the channel is $\Gamma_{out}$ on which we describe the free outflow condition, on the rest of the boundary we prescribe dirichlet values (An parabolic inflow on the left hand side and zero on the upper and lower channel walls). We choose for simplicity $f=0$..\\

@@

In the hp-framework we have a stack of finite elements (a \texttt{hp::FECollection}) given. We assign each cell of the triangulation an fe-index which determines which finite element we use on this cell. The \texttt{ActiveFEIndexSetter} class manages these indices, see there for more information.

>>>>

\subsubsection{General problem description}
\index{stationary PDE}
In this example we consider the same setting as in subsection~\ref{PDE_Stat_Stokes}, the only
difference is that we want to employ the hp-Finite-Elements. So the equation we solve is still 
the stationary incompressible Stokes equation \index{stationary PDE!Stokes equation}. Here,
we use the symmetric stress tensor which has a little consequence when using 
the do-nothing outflow condition. In strong formulation we have
\begin{align} 
-\frac{1}{2}\nabla\cdot (\nabla v + \nabla v^{T}) + \nabla p &= f \\ 
\nabla \cdot v &= 0 \notag
\end{align} 
on the domain $\Omega = [-6,6]\times [0,2]$.  We split $\partial \Omega = \Gamma_D \cup \Gamma_{out}$. The right hand side of the channel is $\Gamma_{out}$ on which we describe the free outflow condition, on the rest of the boundary we prescribe Dirichlet values (An parabolic inflow on the left hand side and zero on the upper and lower channel walls). We choose for simplicity $f=0$..\\

@@

In the hp-framework we have a stack of finite elements (a \texttt{hp::FECollection}) given. We assign each element of the triangulation an fe-index which determines which finite element we use on this element. The \texttt{ActiveFEIndexSetter} class manages these indices, see there for more information.

### 2018-10-17 Examples/PDE/StatPDE/Example8/content.tex
add examples to parallel-dev

Similar to the previous example, we consider the following benchmark problem from plasticity theory:
\begin{equation}
   (\Pi(\sigma(u)),\varepsilon(\varphi)) = (g,\varphi)_{\Gamma_N}.
\end{equation}
Here $\tilde{\Omega}$ is again the quadratic domain with a circular hole around the center cut out. Again, we restrict our actual computational domain $\Omega$ to the upper left quarter of $\tilde{\Omega}$ for reasons of symmetry.\\
We use the symmetric strain tensor $\varepsilon(v) := \frac{1}{2}(\nabla v + \nabla v^T)$, and the symmetric stress tensor $\sigma$ is defined as
\begin{equation*}
   \sigma(v) := 2\mu \varepsilon(v)^D + \kappa \cdot tr (\varepsilon(v)) I
\end{equation*}
where $\tau^D$ is the deviatoric part of a tensor $\tau$, in two dimensions defined as
\begin{equation*}
   \tau^D := \tau - \frac{1}{2} tr(\tau) I.
\end{equation*}
The main difference with respect to the elastic case is the projection operator $\Pi$ in equation $(1)$. It is defined as follows:
\begin{equation*}
\Pi(\tau) = \left\{
            \begin{array}{lr}
            \tau & |\tau^D| \leq \sigma_0\\
            \sigma_0 |\tau^D|^{-1} \tau^D + \frac{1}{2} tr(\tau) I & |\tau^D| > \sigma_0\\ 
            \end{array}
            \right. 
\end{equation*}
In our computations, we choose $\sigma_0 = \sqrt{\frac{2}{3}}\cdot 450$, and the above parameters $\mu$ and $\kappa$ as $\mu = 80193.800283$ resp. $\kappa = 190937.589172$. The corner points of our computational domain are the same as before, and the boundary conditions are not altered, either.\\
The goal of our computations is to detect a subdomain in $\Omega$ where plastic behavior occurs (compare \textit{E. Stein (editor), Error-controlled Adaptive Finite Elements in Solid Mechanics, Wiley (2003), pp. 386 - 389}). This subdomain depends on the right hand side $g$ in equation $(1)$ which we write as $g = \lambda \cdot p$ with $p = 100$ and $\lambda \in [1.5;4.5]$.\\

>>>>

Similar to the previous example, we consider the following benchmark problem from plasticity theory:
\begin{equation}
   (\Pi(\sigma(u)),\varepsilon(\varphi)) = (g,\varphi)_{\Gamma_N}.
\end{equation}
Here $\tilde{\Omega}$ is again the quadratic domain with a circular hole around the center cut out. Again, we restrict our actual computational domain $\Omega$ to the upper left quarter of $\tilde{\Omega}$ for reasons of symmetry.\\
We use the symmetric strain tensor $\varepsilon(v) := \frac{1}{2}(\nabla v + \nabla v^T)$, and the symmetric stress tensor $\sigma$ is defined as
\begin{align*}
   \sigma(v) &:= 2\mu \varepsilon(v)^D + \rho \, tr (\varepsilon(v)) I = 2\mu\varepsilon(v) + \lambda tr(\varepsilon(v)) I,
\end{align*}
where $\tau^D$ is the deviatoric part of a tensor $\tau$, in two dimensions defined as
\begin{equation*}
   \tau^D := \tau - \frac{1}{2} tr(\tau) I.
\end{equation*}
Furthermore, the (standard) Lam\'e parameters are denoted by $\mu$ and
$\lambda$ and which are more conveniently (here and in the code) expressed through
$\rho = \mu + \lambda$ and $\kappa = 2\mu + \lambda$.
The main difference with respect to the elastic case is the projection operator $\Pi$ in equation $(1)$. It is defined as follows:
\begin{equation*}
\Pi(\tau) = \left\{
            \begin{array}{lr}
            \tau & |\tau^D| \leq \sigma_0\\
            \sigma_0 |\tau^D|^{-1} \tau^D + \frac{1}{2} tr(\tau) I & |\tau^D| > \sigma_0\\ 
            \end{array}
            \right. 
\end{equation*}
In our computations, we choose $\sigma_0 = \sqrt{\frac{2}{3}}\cdot 450$, and
the above parameters $\mu, \lambda$ and $\rho$ as $\mu = 80193.800283, \lambda
= 110743.788889, $ and $\rho = 190937.589172$, respectively. The corner points of our computational domain are the same as before, and the boundary conditions are not altered, either.\\
The goal of our computations is to detect a subdomain in $\Omega$ where plastic behavior occurs (compare \textit{E. Stein (editor), Error-controlled Adaptive Finite Elements in Solid Mechanics, Wiley (2003), pp. 386 - 389}). This subdomain depends on the right hand side $g$ in equation $(1)$ which we write as $g = \lambda \cdot p$ with $p = 100$ and $\lambda \in [1.5;4.5]$.\\
